Increased perceptions of autonomy through choice fail to enhance motor skill retention

### 2021-10-31 docs/r0_submission/r0_manuscript.Rmd
update: rename directory and files

---
title             : "Increased perceptions of autonomy through choice fail to enhance motor skill retention"
shorttitle        : "Autonomy-support does not benefit motor learning"
author: 
  - name          : "Laura St. Germain"
    affiliation   : "1"
    corresponding : no    # Define only one corresponding author
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Data curation
      - Formal analysis
      - Investigation - Performed the experiment
      - Methodology
      - Project administration
      - Software - Task Programming
      - Validation
      - Visualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Allison Williams"
    affiliation   : "1"
    role:
      - Investigation - Performed the experiment
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Noura Balbaa"
    affiliation   : "1"
    role:
      - Investigation - Performed the experiment
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Andrew Poskus"
    affiliation   : "1"
    role:
      - Investigation - Performed the experiment
      - Writing - Review & Editing
  - name          : "Olena Leshchyshen"
    affiliation   : "1"
    role:
      - Investigation - Performed the experiment
      - Writing - Review & Editing
  - name          : "Keith R. Lohse"
    affiliation   : "2"
    role:
      - Conceptualization
      - Data curation
      - Formal analysis
      - Methodology
      - Validation
      - Writing - Original Draft Preparation
      - Writing - Review & Editing
  - name          : "Michael J. Carter"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "1280 Main Street West, Ivor Wynne Centre Room 203, McMaster University, Hamilton ON Canada, L8S 4K1"
    email         : "michaelcarter@mcmaster.ca"
    role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
      - Conceptualization
      - Data curation
      - Formal analysis
      - Funding acquisition
      - Methodology
      - Project administration
      - Resources
      - Software - Task Programming
      - Supervision
      - Validation
      - Visualization
      - Writing - Original Draft Preparation
      - Writing - Review & Editing

@@

  We have outlined author contributions using CRediT (Contributor Roles Taxonomy - https://casrai.org/credit/).

abstract: |
  There has been growing research interest in the effects that motivation plays in motor learning, and specifically how autonomy, competence, and social relatedness may directly benefit the learning process. Here, we present a preregistered manipulation of autonomy-support by providing learners with choice during the practice of a speed cup-stacking skill. One group was given control over when a video demonstration was provided and the viewing speed. A yoked control group received an identical demonstration schedule, but without choice (as their schedule was matched to a participant with choice). Critically, we addressed a gap in the literature by adding a yoked group who was explicitly told that they were being denied choice and that their schedule was chosen by another participant. We found no statistically significant learning differences between groups, despite finding evidence that providing choice increased perceived autonomy. Equivalence tests further showed that although the groups were not statistically equivalent, the effect size is likely too small to practically study the effects of autonomy-support through choice in most motor learning labs. These findings add to a growing body of research that questions a causal role of autonomy-support on motor learning, and the robustness of the so-called self-controlled learning advantage.

keywords          : "Motor learning; Pre-registered; Self-controlled; Observation; Equivalence testing"
#wordcount         : "4926"
bibliography      : ["../references.bib", "../r-references.bib"]

@@

csl               : "../apa-single-spaced.csl"
documentclass     : "apa7"
classoption       : "man" # man for manuscript & jou for journal
output            : papaja::apa6_pdf
header-includes   :
  - \pagewiselinenumbers
  - \raggedbottom
  - \leftheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  - \rightheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  #- \hypersetup{draft} # comment out after but right now will prevent error message about some link spanning multiple pages
---

```{r setup, include = FALSE}
library(papaja)
r_refs("r-references.bib")
```

@@

table_1 <- data.frame(c(
  "Self-Controlled",
  "", "",
  "Traditional Yoked",
  "", "",
  "Explicit Yoked", 
  "", ""),
  c(
    "Before each trial, you will be asked whether you wish to watch a modeled demonstration of the task.",
    "If you choose YES, you will then be asked whether you want to watch the video in real-time or in",
    "slow-motion.",

    "Before each trial, you may or may not watch a modeled demonstration of the task based on a pre-",
    "determined schedule. If you observe a model, it might be presented in real-time or in slow-motion",
    "based on the pre-determined schedule.",

    "Before each trial, you may or may not watch a modeled demonstration of the task based on the", 
    "schedule another participant selected. If you observe a model, it might be presented in real-time or in",
    "slow-motion based on what that participant selected."
  )
)

apa_table(table_1, font_size = "scriptsize", col.names = c(" ", "Acquisition instructions"), booktabs = TRUE, span_text_columns = TRUE, caption = "Group specific instructions detailing the observation schedule that would be experienced during Acquisition.")

@@

Participants completed two data collection sessions separated by approximately 24 hours. Session 1 consisted of a pre-test and an acquisition phase. Session 2 consisted of a delayed retention test. At the start of each phase of the experiment, all participants received phase-specific instructions (see Data and code availability section). Group specific instructions were provided prior to the acquisition phase (see Table \@ref(tab:table1)). The instructions appeared on a 22-inch computer monitor (1920x1080 resolution) positioned to the right of the participant. Participants followed along as the instructions were read aloud by the researcher.

@@

Our primary outcome measure was stacking time (i.e., response time) in seconds. Trials recorded as errors $(76/5250 = 1.45\%)$ during data collection were manually removed prior to data analysis. For each participant, pre-test and delayed retention trials were aggregated into one block of five trials and acquisition was aggregated into five blocks of five trials. Significance level was set to 0.05 for all statistical tests. Effect sizes for omnibus tests are reported using generalized eta squared $(\eta^2_{G})$ or eta squared $(\eta^2)$. Post hoc comparisons were conducted using Holm's correction. A Cook's distance of $\geq 1$ was used to identify any influential cases.

@@

```{r fig1, echo = FALSE, fig.cap = "(ref:fig1-caption)", fig.align = "center", out.height = "55%"}
knitr::include_graphics("../../figs/fig1.pdf")
```

@@

table_2 <- data.frame(c(
  "Model viewing frequency",
  " ",
  "Regular speed frequency"
  ),
  c(
    "32.6%",
    " ",
    "33.8%"
  ),
  c(
    "16.9%",
    " ",
    "38.1%"
  ),
  c(
    "12.0%",
    " ",
    "50.0%"
  ),
  c(
    "12.9%",
    " ",
    "34.4%"
  ),
  c(
    "9.3%",
    " ",
    "30.4%"
  )
)

apa_table(table_2, font_size = "scriptsize", col.names = c(" ", "Block 1", "Block 2", "Block 3", "Block 4", "Block 5"), booktabs = TRUE, span_text_columns = TRUE, caption = "The mean frequencies for each acquisition block of requesting to watch the video demonstration and choosing the regular viewing speed.", placement = "t")

@@

```{r fig2, echo = FALSE, fig.cap = "(ref:fig2-caption)", fig.align = "center", out.height = "75%"}
knitr::include_graphics("../../figs/fig2.pdf")

@@

To further probe our null effects [@harms2018] related to our main predictions, two equivalence tests (Self-Controlled versus Traditional Yoked and Traditional Yoked versus Explicit Yoked) on pre-test adjusted retention stacking times were carried out using the two one-sided test procedure [@lakens2018; @schuirmann1987]. To establish our smallest effect size of interest [@lakens2021] we calculated the effect size our design had 33% power to detect [@simonsohn2015], which was $d_{s} = .31$. The Self-Controlled and Traditional Yoked groups were not statistically different, $t(97.99) = -0.830$, $p = 0.409$, and not statistically equivalent, $t(97.99) = 0.720$, $p = 0.237$, given equivalence bounds of $-.355$ and $.355$ (on a raw scale). Similarly, the Traditional Yoked and Explicit Yoked groups were not statistically different, $t(97.99) = 0.262$, $p = 0.794$, and not statistically equivalent, $t(97.99) = -1.288$, $p = 0.100$, given equivalence bounds of $-.355$ and $.355$ (on a raw scale).

@@

It has been argued that self-controlled practice conditions are effective for motor learning because they are autonomy-supportive in nature [@wulf2016]. Yet, this claim has received, at best, modest support in the motor learning literature [e.g., @mckay2020a; but see @barros2019; @carter2017a; @ste-marie2013 for non-support]. Here, we addressed a possible methodological limitation for this lack of support. That is, in previous self-controlled motor learning research, participants in the yoked group are not made aware that they have been denied opportunities to exercise choice over a practice variable such as feedback or watching a modeled demonstration. In the present experiment we introduced a novel yoked group that was explicitly told that the observation schedule---frequency and speed of video---they would receive during practice was actually one created by another participant in the experiment. Contrary to our prediction, the Explicit Yoked group did not report significantly lower perceived autonomy scores than the Traditional Yoked group, which was only informed their observation schedule was predetermined. In line with the "OPTIMAL" theory [@wulf2016], the Self-Controlled group reported significantly higher perceived autonomy-support than both the Traditional and Explicit Yoked groups. However, this boost in perceived autonomy did not lead to enhanced motor performance or learning. These findings from a well-powered experiment are difficult to reconcile with a core pillar of the "OPTIMAL" theory of motor learning.

@@

In sum, we found that being able to control two aspects of one's observation schedule during practice was perceived as autonomy-supportive relative to not having this same control opportunity. However, making participants in an Explicit Yoked group aware they would be denied a control opportunity during practice did not decrease perceptions of autonomy relative to participants in the Traditional Yoked group. Furthermore, we did not find evidence than an autonomy-supportive manipulation had any direct causal effect on learning and we also failed to replicate the typical self-controlled learning advantage. It is worth noting that in some situations autonomy-support in and of itself might be a desired affective outcome [@ste-marie2020]. In such circumstance, providing learners with choice opportunities may be a useful motivationally-protective outcome. However, such a tool should be used judiciously as it is typically motor outcomes in the form of relatively permanent performance changes that is the true goal of any motor learning intervention [@ste-marie2020]. Ultimately, our results add to a growing body of evidence that questions whether autonomy-supportive manipulations directly affect learning. One explanation for this lack of replicability is that the *"true"* effect of such manipulations are much smaller than previously estimated [@mckay2021]. Given the resources required (in terms of sample size) to reliably detect the tiny effects, we encourage motor learning scientists to invest their limited resources carefully---either studying practice conditions that have much bigger effects or devising ways to reduce variability and increase power through clever experimental design [@mcclelland2000].

@@

# Funding
This work was supported by the Natural Sciences and Engineering Research Council (NSERC) of Canada (RGPIN-2018-05589; MJC). LSG was also supported by an NSERC PGS-D scholarship.

# Data and code availability
All material, data, and scripts to reproduce our analyses and figures can be accessed here: https://github.com/cartermaclab/expt_explicit-yoked.

>>>>

---
title             : "Increased perceptions of autonomy through choice fail to enhance motor skill retention"
shorttitle        : "Autonomy-support does not benefit motor learning"
author: 
  - name          : "Laura St. Germain"
    affiliation   : "1"
    corresponding : no    # Define only one corresponding author
    # role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
    #   - Conceptualization
    #   - Data curation
    #   - Formal analysis
    #   - Investigation - Performed the experiment
    #   - Methodology
    #   - Project administration
    #   - Software - Task Programming
    #   - Validation
    #   - Visualization
    #   - Writing - Original Draft Preparation
    #   - Writing - Review & Editing
  - name          : "Allison Williams"
    affiliation   : "1"
    # role:
    #   - Investigation - Performed the experiment
    #   - Writing - Original Draft Preparation
    #   - Writing - Review & Editing
  - name          : "Noura Balbaa"
    affiliation   : "1"
    # role:
    #   - Investigation - Performed the experiment
    #   - Writing - Original Draft Preparation
    #   - Writing - Review & Editing
  - name          : "Andrew Poskus"
    affiliation   : "1"
    # role:
    #   - Investigation - Performed the experiment
    #   - Writing - Review & Editing
  - name          : "Olena Leshchyshen"
    affiliation   : "1"
    # role:
    #   - Investigation - Performed the experiment
    #   - Writing - Review & Editing
  - name          : "Keith R. Lohse"
    affiliation   : "2"
    # role:
    #   - Conceptualization
    #   - Data curation
    #   - Formal analysis
    #   - Methodology
    #   - Validation
    #   - Writing - Original Draft Preparation
    #   - Writing - Review & Editing
  - name          : "Michael J. Carter"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "1280 Main Street West, Ivor Wynne Centre Room 203, McMaster University, Hamilton ON Canada, L8S 4K1"
    email         : "michaelcarter@mcmaster.ca"
    # role:         # Contributorship roles (e.g., CRediT, https://casrai.org/credit/)
    #   - Conceptualization
    #   - Data curation
    #   - Formal analysis
    #   - Funding acquisition
    #   - Methodology
    #   - Project administration
    #   - Resources
    #   - Software - Task Programming
    #   - Supervision
    #   - Validation
    #   - Visualization
    #   - Writing - Original Draft Preparation
    #   - Writing - Review & Editing

@@

  Preregistration: https://aspredicted.org/ze8cj.pdf

  Data, code, and materials: https://github.com/cartermaclab/expt_explicit-yoked

  PsyArXiv preprint: https://doi.org/10.31234/osf.io/s34fg

abstract: |
  There has been growing research interest in the effects that motivation plays in motor learning, and specifically how autonomy, competence, and social relatedness may directly benefit the learning process. Here, we present a preregistered manipulation of autonomy-support by providing learners with choice during the practice of a speed cup-stacking task. One group was given control over when a video demonstration was provided and the viewing speed. A yoked control group received an identical demonstration schedule, but without choice (as their schedule was matched to a participant with choice). Critically, we addressed a gap in the literature by adding a yoked group who was explicitly told that they were being denied choice and that their schedule was chosen by another participant. We found no statistically significant learning differences between groups, despite finding evidence that providing choice increased perceived autonomy. Equivalence tests further showed that although the groups were not statistically equivalent, the effect size is likely too small to practically study the effects of autonomy-support through choice in most motor learning labs. These findings add to a growing body of research that questions a causal role of autonomy-support on motor learning, and the robustness of the so-called self-controlled learning advantage.

  ***Public Significance Statement***  
  The present study suggests that although university-aged students may experience a boost in feelings of autonomy when given choice over features of their learning environment, this boost does not enhance learning of a new motor skill. Additionally, it highlights that coaches and practictioners should be encouraged to incorporate other and more effective practice variables into their training regimes.

keywords          : "Motor learning; Pre-registered; Self-controlled; Observation; Equivalence testing"
wordcount         : "5055"
bibliography      : ["../references.bib", "../r-references.bib"]

@@

csl               : "../apa.csl"
documentclass     : "apa7"
classoption       : "man" # man for manuscript & jou for journal
output            : papaja::apa6_pdf
header-includes   :
  - \usepackage{pdflscape}
  - \pagewiselinenumbers
  - \raggedbottom
  - \leftheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  - \rightheader{\textbf{THIS PREPRINT HAS BEEN SUBMITTED FOR PEER-REVIEW}}
  #- \hypersetup{draft} # comment out after but right now will prevent error message about some link spanning multiple pages
---

```{r setup, include = FALSE}
library(papaja)
library(tidyverse)
library(kableExtra)
# r_refs("../r-references.bib")
```

@@

@@

Participants completed two data collection sessions separated by approximately 24 hours. Session 1 consisted of a pre-test and an acquisition phase. Session 2 consisted of a delayed retention test. At the start of each phase of the experiment, all participants received phase-specific instructions \hyperref[sec:sharing]{(see Data, materials, and code availability section)}. Group specific instructions were provided prior to the acquisition phase (see Table \@ref(tab:table1)). The instructions appeared on a 22-inch computer monitor (1920x1080 resolution) positioned to the right of the participant. Participants followed along as the instructions were read aloud by the researcher. \medskip

table_1 <- tibble(
    Group = c("Self-Controlled",
              "Traditional Yoked",
              "Explicit Yoked"),
    Instructions = c("Before each trial, you will be asked whether you wish to watch a modeled demonstration of the task. If you choose YES, you will then be asked whether you want to watch the video in real-time or in slow-motion.",
                     "Before each trial, you may or may not watch a modeled demonstration of the task based on a pre-determined schedule. If you observe a model, it might be presented in real-time or in slow-motion based on the pre-determined schedule.",
                     "Before each trial, you may or may not watch a modeled demonstration of the task based on the schedule another participant selected. If you observe a model, it might be presented in real-time or in slow-motion based on what that participant selected.")
)

kbl(table_1,
    booktabs = TRUE,
    escape = TRUE,
    linesep = "\\addlinespace",
    caption = "Group specific components from the instructions detailing the observation schedule that would be experienced during Acquisition.",
    col.names = c(" ",
                  "Acquisition instructions")
    ) %>% 
  #landscape() %>% 
  kable_styling(position = "left",
                font_size = 11) %>% 
  column_spec(1, width = "8em") %>% 
  column_spec(2, width = "33em")
```

@@

Our primary outcome measure was stacking time (i.e., response time) in seconds. Trials recorded as errors $(76/5250 = 1.45\%)$ during data collection were manually removed prior to data analysis. For each participant, pre-test and delayed retention trials were aggregated into one block of five trials and acquisition was aggregated into five blocks of five trials. Significance level was set to 0.05 for all statistical tests. Effect sizes for omnibus tests are reported using generalized eta squared $(\eta^2_{G})$ or eta squared $(\eta^2)$. Post hoc comparisons were conducted using Holm's correction. A Cook's distance of $\geq 1$ was used to identify any influential cases and none were identified.

@@

```{r fig1, echo = FALSE, fig.cap = "(ref:fig1-caption)", fig.align = "center", out.height = "51%"}
knitr::include_graphics("../../figs/fig1.pdf")
```

table_2 <- tibble(
    Observation = c("Model viewing frequency",
                    "Regular speed frequency"),
    b1 = c("32.6%",
           "33.8%"),
    b2 = c("16.9%",
           "38.1%"),
    b3 = c("12.0%%",
           "50.0%"),
    b4 = c("12.9%",
           "34.4%"),
    b5 = c("9.3%",
           "30.4%"),
)

kbl(table_2,
    booktabs = TRUE,
    escape = TRUE,
    linesep = "\\addlinespace",
    caption = "The mean frequencies for each Acquisition block of requesting to watch the video demonstration and of those requests, the mean frequency of choosing the regular viewing speed.",
    col.names = c(" ",
                  "Block 1",
                  "Block 2",
                  "Block 3",
                  "Block 4",
                  "Block 5")
    ) %>% 
    kable_styling(position = "left",
                  font_size = 11) %>% 
    column_spec(1, width = "13em") %>% 
    column_spec(2, width = "5em") %>% 
    column_spec(3, width = "5em") %>% 
    column_spec(4, width = "5em") %>% 
    column_spec(5, width = "5em")

@@

@@

```{r fig2, echo = FALSE, fig.cap = "(ref:fig2-caption)", fig.align = "center", out.height = "72%"}
knitr::include_graphics("../../figs/fig2.pdf")

@@

To further probe the null effects [@harms2018] related to our main predictions, two equivalence tests (Self-Controlled versus Traditional Yoked and Traditional Yoked versus Explicit Yoked) on pre-test adjusted retention stacking times were carried out using the two one-sided test procedure [@lakens2018; @schuirmann1987]. To establish our smallest effect size of interest [@lakens2021] we calculated the effect size our design had 33% power to detect [@simonsohn2015], which was $d_{s} = .31$. The Self-Controlled and Traditional Yoked groups were not statistically different, $t(97.99) = -0.830$, $p = 0.409$, and not statistically equivalent, $t(97.99) = 0.720$, $p = 0.237$, given equivalence bounds of $-.355$ and $.355$ (on a raw scale). Similarly, the Traditional Yoked and Explicit Yoked groups were not statistically different, $t(97.99) = 0.262$, $p = 0.794$, and not statistically equivalent, $t(97.99) = -1.288$, $p = 0.100$, given equivalence bounds of $-.355$ and $.355$ (on a raw scale).

@@

It has been argued that self-controlled practice conditions are effective for motor learning because they are autonomy-supportive in nature [@wulf2016]. Yet, this claim has received, at best, modest support in the motor learning literature [e.g., @mckay2020a; but see @barros2019; @carter2017a; @ste-marie2013 for non-support]. Here, we addressed a possible methodological limitation for this lack of support. That is, in previous self-controlled motor learning research, participants in the yoked group are not made aware that they have been denied opportunities to exercise choice over a practice variable such as feedback or watching a modeled demonstration. In the present experiment we introduced a novel yoked group that was explicitly told that the observation schedule---frequency and speed of video---they would receive during practice was actually one created by another participant in the experiment. Contrary to our prediction, the Explicit Yoked group did not report significantly lower perceived autonomy scores than the Traditional Yoked group, which was only informed their observation schedule was predetermined. In line with the "OPTIMAL" theory [@wulf2016], the Self-Controlled group reported significantly higher perceived autonomy-support than both the Traditional and Explicit Yoked groups. However, this boost in perceived autonomy did not lead to enhanced motor performance or learning. These findings from a preregistered experiment are difficult to reconcile with a core pillar of the "OPTIMAL" theory of motor learning.

@@

In sum, we found that being able to control two aspects of one's observation schedule during practice was perceived as autonomy-supportive relative to not having this same control opportunity. However, making participants in an Explicit Yoked group aware they would be denied a control opportunity during practice did not decrease perceptions of autonomy relative to participants in the Traditional Yoked group. Furthermore, we did not find evidence than an autonomy-supportive manipulation had any direct causal effect on learning and we also failed to replicate the typical self-controlled learning advantage. It is worth noting that in some situations autonomy-support in and of itself might be a desired affective outcome [@ste-marie2020]. In such circumstances, providing learners with choice opportunities may be a useful motivationally-protective outcome. However, such a tool should be used judiciously as it is typically motor outcomes in the form of relatively permanent performance changes that is the true goal of any motor learning intervention [@ste-marie2020]. Ultimately, our results add to a growing body of evidence that questions whether autonomy-supportive manipulations directly affect motor learning. One explanation for this lack of replicability is that the *"true"* effect of such manipulations are much smaller than previously estimated [@mckay2021]. Given the resources required (in terms of sample size) to reliably detect these tiny effects, we encourage motor learning scientists to invest their limited resources carefully---either studying practice conditions that have much bigger effects or devising ways to reduce variability and increase power through clever experimental design [@mcclelland2000].

\vfill

# Contributions using CrediT (https://casrai.org/credit/)
Conceptualization (LSG, KRL, MJC); Data curation (LSG, KRL, MJC); Formal analysis (LSG, KRL, MJC); Funding acquisition (MJC); Investigation - Performed Experiment (LSG, AW, NB, AP, OL); Methodology (LSG, KRL, MJC); Project administration (LSG, MJC); Resources (MJC); Software - Task Programming (LSG, MJC); Supervision (MJC); Validation (LSG, KRL, MJC); Visualization (LSG, MJC); Writing - Original Draft Preparation (LSG, AW, NB, KRL, MJC); Writing - Review & Editing (LSG, AW, NB, AP, OL, KRL, MJC)

# Data, materials, and code availability
\label{sec:sharing}
All material, data, and scripts to reproduce our analyses and figures can be accessed here: https://github.com/cartermaclab/expt_explicit-yoked. The preregistration can be accessed here: https://aspredicted.org/ze8cj.pdf.

@@

# Funding
This work was supported by the Natural Sciences and Engineering Research Council (NSERC) of Canada (RGPIN-2018-05589; MJC). LSG was supported by an NSERC Postgraduate Scholarship. AW and AP were supported by NSERC Undergraduate Student Research Awards.
